---
title: "Sample Taxi Report"
author: "Hamed Bastan-Hagh"
output: 
  html_document: 
    highlight: textmate
    theme: united
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(ggplot2)
theme_set(theme_light())
```

This is a simple report for demo purposes. The data wrangling has mostly been done in Python, here we read in those data and generate some plots.

```{r load-data}
d_weekly <- fread(here::here("data/processed/weekly_summary.csv"))
d_hourly <- fread(here::here("data/processed/time_of_day_summary.csv"))
```

We can view some trends over time, for example the average cost of a yellow taxi ride over the period.

```{r cost-weekly}
d_weekly |> 
  ggplot(aes(week_start, avg_cost)) + 
  geom_smooth(
    formula = y ~ x,
    method = "loess",
    se = FALSE,
    colour = "grey30"
  ) +
  geom_line(colour = "yellow3") + 
  geom_point(size = 0.7) +
  scale_y_continuous(labels = scales::label_currency()) +
  labs(
    title = "Average yellow taxi fares 2018-2022",
    subtitle = "Total trip costs averaged by week",
    caption = "Grey line is LOESS curve",
    x = NULL,
    y = NULL
  )
```

The sharp dip in 2020 hints at Covid, which we can see much more clearly in the trip count.

```{r trips-weekly}
d_weekly |> 
  ggplot(aes(week_start, trips)) + 
  geom_smooth(
    formula = y ~ x,
    method = "loess",
    se = FALSE,
    colour = "grey30"
  ) +
  geom_line(colour = "yellow3") + 
  geom_point(size = 0.7) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Yellow taxi trip counts 2018-2022",
    subtitle = "Total trips by week",
    caption = "Grey line is LOESS curve",
    x = NULL,
    y = NULL
  )
```

Not only is there a sharp dip in 2020, but over time there are far fewer yellow taxi rides. Presumably this is down to people switching to Uber.

We can also consider patterns by the time of day: how do average fares differ at different times of day?

```{r cost-hourly}
d_hourly[
  ,
  .(avg_cost = weighted.mean(avg_cost, trips)),
  by = .(year = factor(year), hour)
] |> 
  ggplot() +
  geom_area(
    aes(x = hour, y = avg_cost),
    colour = "black",
    fill = "yellow",
    alpha = 0.7
  ) +
  scale_x_continuous(
    breaks = seq(3, 21, by = 6),
    labels = c("3AM", "9AM", "3PM", "9PM")
  ) +
  scale_y_continuous(labels = scales::label_currency()) +
  facet_wrap(~ year) +
  labs(
    title = "Average yellow taxi fares 2018-2022",
    subtitle = "Total trip costs averaged by time of day",
    x = NULL,
    y = NULL
  )
```

```{r trips-hourly}
d_hourly[
  ,
  .(trips = sum(trips) / 1e6),
  by = .(year = factor(year), hour)
] |> 
  ggplot() +
  geom_area(
    aes(x = hour, y = trips),
    colour = "black",
    fill = "yellow",
    alpha = 0.7
  ) +
  scale_x_continuous(
    breaks = seq(3, 21, by = 6),
    labels = c("3AM", "9AM", "3PM", "9PM")
  ) +
  scale_y_continuous(labels = scales::label_number(suffix = "M")) +
  facet_wrap(~ year) +
  labs(
    title = "Yellow taxi trip counts 2018-2022",
    subtitle = "Total trips by time of day in millions",
    x = NULL,
    y = NULL
  )
```

